### 2023-04-29 main/chapter_experiment.tex
incorporate Val's round 2 comments

\section{The Large Hadron Collider}
\label{sec:atlas_lhc}
% The LHC
% largest piece
The Large Hadron Collider (LHC) is the largest and most powerful piece of
CERN's extensive particle accelerator complex.
It lives in a $26.7\,\mathrm{km}$ (circumference) tunnel, which is buried
around $100\,\mathrm{m}$ under France and Switzerland at CERN, Geneva.
It and comprises two rings that contain and accelerate counter-rotating bunches
of particles.
These accelerated bunches are brought into near-head-on collisions at
the four large detector experiments:
\atlas~\cite{atlas2008experiment},
\lhcb~\cite{lhcb2008experiment},
\cms~\cite{cms2008experiment}, and
\alice~\cite{alice2008experiment}.
Although proton-proton collisions are the LHC's primary subject, it is also
operated for heavy ion collisions including lead ($^{208}\mathrm{Pb}^{82+}$)
and oxygen ($^{16}\mathrm{O}^{4+}$)~\cite{
Brewer:2021kiv,
lhc2008machine,
lhcDesignreport2004
}.
Heavy ion collisions are primarily studied by the \alice\ experiments, but are
also analysed by the general-purpose detectors \atlas\ and \cms.
Since proton collision data are the topic of this thesis, this section focuses
on proton operations.

% does not stand alone
The LHC does not stand alone.
It implements the final boosts and collisions of particles, but requires
earlier stages of CERN's systems to inject the protons at suitably high
energies for it to continue their acceleration.
In brief, the accelerator chain flows as follows:
% Accelerator complex
\begin{enumerate}
% where the protons come from
\item[0.] The protons come from hydrogen gas, which is initially ionised into a
plasma of protons and $e^-$ ions in the strong electric fields of a
`duoplasmotron' ion source, in which the protons are initially accelerated by
electric fields towards a cathode~\cite{lhc50years2013}.
% linac2
\item After ionisation, a linear accelerator named Linac2 provides the first
major proton acceleration in the chain, reaching $50\,\eV[M]$
energies~\cite{lhc50years2013}.
% PSB
\item Secondly, the Proton Synchrotron Booster (PSB) accelerates these protons
to $1.4\,\eV[G]$ energies.
The PSB comprises four rings with $160\,\mathrm{m}$
circumference~\cite{linac4protonsource,lhc50years2013}.
% PS
\item Thirdly, protons from the PSB are accelerated by the
$628\,\mathrm{m}$-circumference
Proton Synchrotron (PS) to $25\,\eV[G]$ energies~\cite{lhc2000psInjector}.
% SPS
\item Fourthly, protons from the PS are further boosted by the
$6.9\,\mathrm{km}$-circumference
Super Proton Synchrotron (SPS) to $450\,\eV[G]$
energies~\cite{synchroton:445034}.
% LHC
\item Finally, protons from the SPS are injected the LHC, which finishes
their acceleration from $250\,\eV[G]$ to $6.5\,\eV[T]$, achieving the
world-leading $\sqrt{s} = 13\,\eV[T]$ proton-proton collisions that are studied
in this thesis.
\end{enumerate}
This sequence is a subset of CERN's full accelerator complex,
illustrated in Figure~\ref{fig:experiment_cern_complex}~\cite{Haffner:1621894},
which also supports various other engineering and research experiments.
Each of these accelerators (Linac2, PSB, PS, and SPS) was upgraded from its
previous operations to achieve the performance demanded by designs for the
LHC's high energy and luminosity.

@@

Radiofrequency (RF) cavities are the technology used to accelerate protons in
the LHC.
They operate by setting up electron beams to oscillate at a frequency that
resonates with a specially designed cavity, this producing strong oscillating
electromagnetic fields.
These electromagnetic fields oscillate at $400\,\mathrm{MHz}$ in the LHC, and
must be carefully tuned jointly with the incident proton beams to align in
their timing.
When perfectly timed, the RF cavity will accelerate protons that are less
energetic that the target beam energy, and decelerate any that are more
energetic.
This achieves a dynamically stable acceleration.
The LHC uses sixteen superconducting RF cavities, which are contained in four
separate cryostatic modules~\cite{lhcDesignreport2004}.

% Magnets
Magnets are used to steer and focus the LHC's proton beams.
% - dipole for steering
Dipole magnets conduct steering by creating constant magnetic fields, with
north and south magnetic poles on diametrically-opposite sides of the beam.
These dipole fields steer the beam by applying constant Lorentz forces
perpendicular to the beam's motion.
% - quadrupole for focussing (pairs)
Quadrupole magnets focus the beam to counteract its natural dispersion, with
alternating magnetic poles placed after each $90^\circ$ rotation around the
beam.
A quadrupole magnet is focussing in one axis, and defocussing in a
perpendicular axis.
However, alternating pairs of quadrupole magnets can achieve
a total focussing effect; their effects are not linear.

@@

% size of beam
By its peak luminosity design specifications, each bunch of an LHC beam
should contain $1.15\times10^{11}$ protons.
The ring can contain up to $2808$ bunches.
After acceleration to collision energy, each bunch has a root-mean-square
(RMS, standard-deviation) length of $7.55\,\mathrm{cm}$.
The beam has a transverse RMS width of $16.7\,\mathrm{\mu m}$ at
interaction points 1 and 5 (\atlas\ and \cms),
and of $70.9\,\mathrm{\mu m}$ at
interaction points 2 and 8 (\alice\ and \lhcb)%
~\cite{lhcDesignreport2004}.

% luminosity and its uncertainty
The luminosity delivered to each detector varies with its local beam controls.
For example, beams at the \lhcb\ and \alice\ interaction points are
deliberately offset to reduce data rates and `pile-up' backgrounds from
multiple interactions~\cite{Follin:2014nva}
The peak instantaneous luminosity delivered by the LHC to \atlas\ in Run~2 was
$1.9\times10^{34}\,\mathrm{cm}^{-2}\mathrm{s}^{-1}$ in 2018;
see Table~\ref{tab:experiment_run2_state} for more details here.
% Van Der Meer scans
Luminosity is measured in \atlas\ during dedicated low-luminosity runs of the
LHC, using the ``van der Meer'' method, which measures interaction rates while
scanning the beams across each other in transverse
space~\cite{vanderMeer:296752,ATLAS:2022hro}.
These \atlas\ luminosity measurements primarily rely on data from the
LUCID2 Cherenkov detector, which comprises two parts that are offset from the
interaction point by about $\pm 17\,\mathrm{m}$ along the beam-pipe~\cite{
SBARRA2019152,
Avoni_2018
}.
The relative uncertainty on \atlas's Run~2 integrated luminosity is $\pm1.7\%$.

@@

% pile-up
Pile-up interactions cause tracks and other detector hits, which are noise
atop our measurements of the primary vertex.
If not removed, these pile-up signals can skew measurements and create an
illusion missing transverse momentum.
Although precise measurements can help to remove this noise, it reduces the
information content of our data and cannot be cleaned entirely.
In Run~2 data, the number of pile-up vertices per bunch crossing varies around
several tens of vertices, as displayed in
Figure~\ref{fig:experiment_run_2_mu}~\cite{atlas_public_luminosity}.
Note that the variability in the mean pile-up count arises from differing beam
conditions over run-time; for example, the instantaneous luminosity can only
decrease during each LHC beam fill, and the expected pile-up rate decreases
with it.

@@

Our search paper~\cite{atlas2022searches} contains concise and standard
descriptions of the \atlas\ detector and its relevant components.
Through its evolution in many past \atlas\ publications, these descriptions
have become so highly refined that attempts to improve them myself would
probably be in vain.
Instead, this section quotes our published text with explanatory commentary.
\begin{quote}
``%
The ATLAS detector~\cite{atlas2008experiment} is a multipurpose particle
detector with a forward--backward symmetric cylindrical geometry and a near
$4\pi$ coverage in solid angle.%
\footnote{%
``ATLAS uses a right-handed coordinate system with its origin at the nominal
interaction point (IP) in the centre of the detector and the $z$-axis along the
beam pipe.
The $x$-axis points from the IP to the centre of the LHC ring, and the $y$-axis
points upwards.
Cylindrical coordinates $(r,\phi)$ are used in the transverse plane, $\phi$
being the azimuthal angle around the $z$-axis.
The pseudorapidity is defined in terms of the polar angle $\theta$ as
$\eta = -\ln \tan(\theta/2)$.
Angular distance is measured in units of
$\Delta R \equiv \sqrt{(\Delta\eta)^{2} + (\Delta\phi)^{2}}$.%
''\footnotemark~\cite{atlas2022searches}%
}
It consists of an inner tracking detector surrounded by a thin superconducting
solenoid providing a $2\,\mathrm{T}$ axial magnetic field, electromagnetic and
hadronic calorimeters, and a muon spectrometer.%
''~\cite{atlas2022searches}
\end{quote}
To illustrate this summary, Figure~\ref{fig:atlas_cutaway} shows a vertical slice
of \atlas.
Each subsystem is described in more detail in a subsection below:
the inner tracking detector in Section~\ref{sec:atlas_inner},
the electromagnetic and hadronic calorimeters in
Section~\ref{sec:atlas_calo},
and
the muon spectrometer in Section~\ref{sec:atlas_muon}.
These detectors' resolutions to relevant physical objects is described in
Section~\label{sec:atlas_object_resolutions}.
The final connection to data analysed in this thesis is made through the
trigger system, which is described in Section~\ref{sec:atlas_trigger},
and details of the data themselves are given in
Section~\ref{sec:atlas_data}.

@@

Energy losses from high-energy electrons are dominated by Bremsstrahlung
radiations, in which the electron radiates a photon and recoils from its
momentum.
Electron reconstruction algorithms efficiently recover this lost Bremsstrahlung
momentum by associating nearby photon calorimeter clusters back to their source
electron many cases~\cite{atlas1994proposal, atlas1999design1}.

\paragraph{Jets.}
Although charged hadrons are tracked in the inner detector, the total energy of
jets also depends vitally on neutral hadrons, which are only observed through
their hadronic interactions in the calorimeter system.
With less influence from tracking, calorimetry noise is the dominant
contribution to the experimental resolution on hadronic jets.
However, other noise contributions arising from electrical signals and pile-up
scale in proportion in $1/\pt$.
one can therefore again approximate \atlas's relative jet energy resolution as
\begin{equation}
\label{eqn:experiment_jet_resolution}
\frac{\sigma(\pt)}{\pt}
\approx
\frac{b_1}{\sqrt{\pt}}
\oplus
\frac{b_2}{\pt}
\oplus
b_3
,
\end{equation}
the same functional form as for electrons in
Equation~\ref{eqn:experiment_electron_resolution},
with
$b_1 \approx 1.0\,\sqrt{\eV[G]}$
$b_2 \approx 0.5\,\eV[G]$, and
$b_3 \approx 0.03$.
With these parameters, formula approximately describes the measured relative
jet resolutions of
$\frac{\sigma(\pt)}{\pt} \approx 20\%$ at $\pt = 30\,\eV[G]$,
$\frac{\sigma(\pt)}{\pt} \approx 8\%$ at $\pt = 200\,\eV[G]$, and
$\frac{\sigma(\pt)}{\pt} \approx 4\%$ at
$\pt = 2\,\eV[T]$~\cite{atlas2021jetresolution}.

@@

which are highlighted in bold.
Table adapted from~\cite{ATL-DAQ-PUB-2019-001}.

@@

@@

\begin{sidewaystable}[tp]
\centering
\begin{tabular}{lcccc}
Observable                                                        & 2015         & 2016         & 2017                & 2018         \\
\hline
Maximum number of colliding bunch-pairs                           & $2232$       & $2208$       & $2544$ ($1909$)     & $2544$       \\
Bunch spacing ($\mathrm{ns}$)                                     & $25$         & $25$         & $25$ (variable)     & $25$         \\
Typical number of protons per bunch ($\times 10^{11}$)            & $1.1$        & $1.1$        & $1.1$ ($1.2$)       & $1.1$        \\
Peak luminosity ($\times 10^{33}\mathrm{cm}^{-2}\mathrm{s}^{-1}$) & $5$          & $13$         & $16$                & $19$         \\
Peak number of inelastic interactions per crossing                & $\approx 16$ & $\approx 41$ & $\approx 45$ ($60$) & $\approx 55$ \\
Luminosity-weighted average inelastic interactions per crossing   & $13$         & $25$         & $38$                & $36$         \\
Integrated luminosity delivered ($\mathrm{fb}^{-1}$)              & $4$          & $39$         & $50.6$              & $63.8$
\end{tabular}
\caption[
LHC data rates and conditions by year in Run~2
]{%
LHC data rates and conditions by year in Run~2.
This table is adapted from~\cite{ATLAS:2022hro}.
Bracketed numbers under 2017 refer to an alternative `8b4e' LHC operating
condition which alternated in a pattern of eight bunches with the usual
$25\,\mathrm{ns}$ separation followed by a gap of four empty bunch slots.
Alongside a primary interaction, the inelastic interactions in a crossing are
pile-up events and add noise to our measurements.
}
\label{tab:experiment_run2_state}
\end{sidewaystable}

>>>>

\section{The Large Hadron Collider}
\label{sec:atlas_lhc}
% The LHC
% largest piece
The Large Hadron Collider (LHC) is the largest and most powerful piece of
CERN's extensive particle accelerator complex.
It lives in a $26.7\,\mathrm{km}$ (circumference) tunnel, which is buried
around $100\,\mathrm{m}$ under France and Switzerland at CERN, Geneva.
It comprises two rings that contain and accelerate counter-rotating bunches
of particles.
These accelerated bunches are brought into near-head-on collisions at
the four large detector experiments:
\atlas~\cite{atlas2008experiment},
\lhcb~\cite{lhcb2008experiment},
\cms~\cite{cms2008experiment}, and
\alice~\cite{alice2008experiment}.
Although proton-proton collisions are the LHC's primary subject, it is also
operated for heavy ion collisions including lead ($^{208}\mathrm{Pb}^{82+}$)
and oxygen ($^{16}\mathrm{O}^{4+}$)~\cite{
Brewer:2021kiv,
lhc2008machine,
lhcDesignreport2004
}.
Heavy ion collisions are primarily studied by the \alice\ experiments, but are
also analysed by the general-purpose detectors \atlas\ and \cms,
as well as \lhcb.
Since proton collision data are the topic of this thesis, this section focuses
on proton operations.

% does not stand alone
The LHC does not stand alone.
It implements the final boosts and collisions of particles, but requires
earlier stages of CERN's systems to inject the protons at suitably high
energies for it to continue their acceleration.
In brief, the accelerator chain flows as follows:
% Accelerator complex
\begin{enumerate}
% where the protons come from
\item[0.] The protons come from hydrogen gas, which is initially ionised into a
plasma of protons and $e^-$ ions in the strong electric fields of a
`duoplasmotron' ion source, in which the protons are initially accelerated by
electric fields towards a cathode~\cite{lhc50years2013}.
% linac2
\item After ionisation, a linear accelerator named Linac2 provides the first
major proton acceleration in the chain, reaching $50\,\eV[M]$
energies~\cite{lhc50years2013}.
% PSB
\item Secondly, the Proton Synchrotron Booster (PSB) accelerates these protons
to $1.4\,\eV[G]$ energies.
The PSB comprises four rings with $160\,\mathrm{m}$
circumference~\cite{linac4protonsource,lhc50years2013}.
% PS
\item Thirdly, protons from the PSB are accelerated by the
$628\,\mathrm{m}$-circumference
Proton Synchrotron (PS) to $25\,\eV[G]$ energies~\cite{lhc2000psInjector}.
% SPS
\item Fourthly, protons from the PS are further boosted by the
$6.9\,\mathrm{km}$-circumference
Super Proton Synchrotron (SPS) to $450\,\eV[G]$
energies~\cite{synchroton:445034}.
% LHC
\item Finally, protons from the SPS are injected into the LHC, which finishes
their acceleration from $250\,\eV[G]$ to $6.5\,\eV[T]$, achieving the
world-leading $\sqrt{s} = 13\,\eV[T]$ proton-proton collisions that are studied
in this thesis.
\end{enumerate}
This sequence is a subset of CERN's full accelerator complex,
illustrated in Figure~\ref{fig:experiment_cern_complex}~\cite{Haffner:1621894},
which also supports various other engineering and research experiments.
Each of these accelerators (Linac2, PSB, PS, and SPS) was upgraded from its
previous operations to achieve the performance demanded by designs for the
LHC's high energy and luminosity.

@@

Radiofrequency (RF) cavities are the technology used to accelerate protons in
the LHC.
They operate by setting up electron beams to oscillate at a frequency that
resonates with a specially designed cavity, thus producing strong oscillating
electromagnetic fields.
These electromagnetic fields oscillate at $400\,\mathrm{MHz}$ in the LHC, and
must be carefully tuned jointly with the incident proton beams to align in
their timing.
When perfectly timed, the RF cavity will accelerate protons that are less
energetic that the target beam energy, and decelerate any that are more
energetic.
This achieves a dynamically stable acceleration.
The LHC uses sixteen superconducting RF cavities, which are contained in four
separate cryostatic modules~\cite{lhcDesignreport2004}.

% Magnets
Magnets are used to steer and focus the LHC's proton beams.
% - dipole for steering
Dipole magnets conduct steering by creating constant magnetic fields, with
north and south magnetic poles on diametrically-opposite sides of the beam.
These dipole fields steer the beam by applying constant Lorentz forces
perpendicular to the beam's motion.
% - quadrupole for focussing (pairs)
Quadrupole magnets focus the beam to counteract its natural dispersion, with
alternating magnetic poles placed after each $90^\circ$ rotation around the
beam.
A quadrupole magnet is focussing in one axis, and defocussing in a
perpendicular axis.
However, alternating pairs of quadrupole magnets can achieve
a total focussing effect; their effects are not linear.
% "Give number of magnets" - internal examiner
The LHC includes
$1232$ dipole magnets
and $4800$ higher-pole `corrector' magnets~\cite{lhc2008machine}.

@@

% size of beam
At its peak luminosity by design specifications, each bunch of an LHC beam
should contain $1.15\times10^{11}$ protons.
The ring can contain up to $2808$ bunches.
After acceleration to collision energy, each bunch has a root-mean-square
(RMS, standard-deviation) length of $7.55\,\mathrm{cm}$.
The beam has a transverse RMS width of $16.7\,\mathrm{\mu m}$ at
interaction points 1 and 5 (\atlas\ and \cms),
and of $70.9\,\mathrm{\mu m}$ at
interaction points 2 and 8 (\alice\ and \lhcb)%
~\cite{lhcDesignreport2004}.

% luminosity and its uncertainty
The luminosity delivered to each detector varies with its local beam controls.
For example, beams at the \lhcb\ and \alice\ interaction points are
deliberately offset to reduce data rates and `pile-up' backgrounds from
multiple interactions~\cite{Follin:2014nva}
The peak instantaneous luminosity delivered by the LHC to \atlas\ in Run~2 was
$1.9\times10^{34}\,\mathrm{cm}^{-2}\mathrm{s}^{-1}$ in 2018;
see Table~\ref{tab:experiment_run2_state} for more details here.
% Van Der Meer scans
Luminosity is measured in \atlas\ during dedicated low-luminosity runs of the
LHC, using the ``van der Meer'' method, which measures interaction rates while
scanning the beams across each other in transverse
space~\cite{vanderMeer:296752,ATLAS:2022hro}.
These \atlas\ luminosity measurements primarily rely on data from the
LUCID2 Cherenkov detector, which comprises two parts that are offset from the
interaction point by about $\pm 17\,\mathrm{m}$ along the beam-pipe~\cite{
SBARRA2019152,
Avoni_2018
}.
The relative uncertainty on \atlas's Run~2 integrated luminosity is
$\pm1.7\%$~\cite{atlas2022searches}.

@@

% pile-up
Pile-up interactions cause tracks and other detector hits, which are noise
atop our measurements of the primary vertex.
If not removed, these pile-up signals can skew measurements and create an
illusion of missing transverse momentum.
Although precise measurements can help to remove this noise, it reduces the
information content of our data and cannot be cleaned entirely.
In Run~2 data, the number of pile-up vertices per bunch crossing varies around
several tens of vertices, as displayed in
Figure~\ref{fig:experiment_run_2_mu}~\cite{atlas_public_luminosity}.
Note that the variability in the mean pile-up count arises from differing beam
conditions over run-time; for example, the instantaneous luminosity can only
decrease during each LHC beam fill, and the expected pile-up rate decreases
with it.

@@

Our search paper~\cite{atlas2022searches} contains concise and standard
descriptions of the \atlas\ detector and its relevant components.
Through its evolution in many past \atlas\ publications, these descriptions
have become so highly refined that attempts to improve them myself would
probably be in vain.
Instead, this section quotes our published text with explanatory commentary.
\begin{quote}
``%
The ATLAS detector~\cite{atlas2008experiment} is a multipurpose particle
detector with a forward--backward symmetric cylindrical geometry and a near
$4\pi$ coverage in solid angle.%
\footnote{%
``ATLAS uses a right-handed coordinate system with its origin at the nominal
interaction point (IP) in the centre of the detector and the $z$-axis along the
beam pipe.
The $x$-axis points from the IP to the centre of the LHC ring, and the $y$-axis
points upwards.
Cylindrical coordinates $(r,\phi)$ are used in the transverse plane, $\phi$
being the azimuthal angle around the $z$-axis.
The pseudorapidity is defined in terms of the polar angle $\theta$ as
$\eta = -\ln \tan(\theta/2)$.
Angular distance is measured in units of
$\Delta R \equiv \sqrt{(\Delta\eta)^{2} + (\Delta\phi)^{2}}$.%
''\footnotemark~\cite{atlas2022searches}%
}
It consists of an inner tracking detector surrounded by a thin superconducting
solenoid providing a $2\,\mathrm{T}$ axial magnetic field, electromagnetic and
hadronic calorimeters, and a muon spectrometer.%
''~\cite{atlas2022searches}
\end{quote}
To illustrate this summary, Figure~\ref{fig:atlas_cutaway} shows a vertical slice
of \atlas.
Each subsystem is described in more detail in a subsection below:
the inner tracking detector in Section~\ref{sec:atlas_inner},
the electromagnetic and hadronic calorimeters in
Section~\ref{sec:atlas_calo},
and
the muon spectrometer in Section~\ref{sec:atlas_muon}.
These detectors' resolutions to relevant physical objects is described in
Section~\ref{sec:atlas_object_resolutions}.
The final connection to data analysed in this thesis is made through the
trigger system, which is described in Section~\ref{sec:atlas_trigger},
and details of the data themselves are given in
Section~\ref{sec:atlas_data}.

@@

Energy losses from high-energy electrons are dominated by Bremsstrahlung
radiations, in which the electron radiates a photon and recoils from its
momentum.
Electron reconstruction algorithms efficiently recover this lost Bremsstrahlung
momentum by associating nearby photon-induced calorimeter clusters back to
their source electron many cases~\cite{atlas1994proposal, atlas1999design1}.

\paragraph{Jets.}
Although charged hadrons are tracked in the inner detector, the total energy of
jets also depends vitally on neutral hadrons, which are only observed through
their hadronic interactions in the calorimeter system.
With less influence from tracking, calorimetry noise is the dominant
contribution to the experimental resolution on hadronic jets.
However, other noise contributions arising from electrical signals and pile-up
scale in proportion in $1/\pt$.
one can therefore again approximate \atlas's relative jet energy resolution as
\begin{equation}
\label{eqn:experiment_jet_resolution}
\frac{\sigma(\pt)}{\pt}
\approx
\frac{b_1}{\sqrt{\pt}}
\oplus
\frac{b_2}{\pt}
\oplus
b_3
,
\end{equation}
the same functional form as for electrons in
Equation~\ref{eqn:experiment_electron_resolution},
with
$b_1 \approx 1.0\,\sqrt{\eV[G]}$
$b_2 \approx 0.5\,\eV[G]$, and
$b_3 \approx 0.03$.
With these parameters, this formula approximately describes the measured
relative jet resolutions of
$\frac{\sigma(\pt)}{\pt} \approx 20\%$ at $\pt = 30\,\eV[G]$,
$\frac{\sigma(\pt)}{\pt} \approx 8\%$ at $\pt = 200\,\eV[G]$, and
$\frac{\sigma(\pt)}{\pt} \approx 4\%$ at
$\pt = 2\,\eV[T]$~\cite{atlas2021jetresolution}.

@@

which are highlighted in \textbf{bold}.

@@

This table is adapted from~\cite{ATL-DAQ-PUB-2019-001}.

@@

\begin{sidewaystable}[tp]
\centering
\begin{tabular}{lcccc}
Observable                                                        & 2015         & 2016         & 2017                & 2018         \\
\hline
Maximum number of colliding bunch-pairs                           & $2232$       & $2208$       & $2544$ ($1909$)     & $2544$       \\
Bunch spacing ($\mathrm{ns}$)                                     & $25$         & $25$         & $25$ (variable)     & $25$         \\
Typical number of protons per bunch ($\times 10^{11}$)            & $1.1$        & $1.1$        & $1.1$ ($1.2$)       & $1.1$        \\
Peak luminosity ($\times 10^{33}\mathrm{cm}^{-2}\mathrm{s}^{-1}$) & $5$          & $13$         & $16$                & $19$         \\
Peak number of inelastic interactions per crossing                & $\approx 16$ & $\approx 41$ & $\approx 45$ ($60$) & $\approx 55$ \\
Luminosity-weighted average inelastic interactions per crossing   & $13$         & $25$         & $38$                & $36$         \\
Integrated luminosity delivered ($\mathrm{fb}^{-1}$)              & $4$          & $39$         & $50.6$              & $63.8$
\end{tabular}
\caption[
LHC data rates and conditions by year in Run~2
]{%
LHC data rates and conditions by year in Run~2.
This table is adapted from~\cite{ATLAS:2022hro}.
Bracketed numbers under 2017 refer to an alternative `8b4e' LHC operating
condition, which alternated in a pattern of eight bunches with the usual
$25\,\mathrm{ns}$ separation followed by a gap of four empty bunch slots.
Alongside a primary interaction, the inelastic interactions in a crossing are
pile-up events and add noise to our measurements.
}
\label{tab:experiment_run2_state}
\end{sidewaystable}

### 2023-04-29 main/chapter_searches.tex
incorporate Val's round 2 comments

Despite their troublesome nomenclature, \heplikelihood\ models are extremely
useful and effective for describing the predictions of our hypotheses and
extracting interpretable results.
A general \heplikelihood\ model is a function $H(x)$ that takes input
parameters $x$ and outputs two things:
expected yields $y(x)$ to predict Poisson data,
and an objective value $E$ to be minimized (in a `fit') to choose a best
prediction.
That is,
\begin{equation}
H(x) = [\,y(x), \,E(x)\,]
.
\end{equation}
The minimization objective $E$ is the summed negative $\log$ \heplikelihood\
for all actual and auxiliary data.
We choose the letter $E$ for \textbf{energy}, to highlight the analogy with
statistical physics and the Boltzmann factor
$f(x) \propto e^{-E(x)/(k_B T)}$,
with which a thermodynamic system collapses to its minimum energy at cold
temperatures~\cite{
skilling2017david,
lecun-06,
pmlr-v2-ranzato07a
}.
At temperature $1/k_B$, this can induce a probability distribution
relative to an appropriate measure~\cite{
cranmer2012histfactory,
skilling2017david
},
but in my experience with \heplikelihood\ models such distributions often
coincide with neither intuition nor the best fit value.%
\footnote{%
One can reproduce this experience by sampling a public \pyhf\ \heplikelihood\
model
(for example, try SR-HM(disc.) of $1\ell bb$~\cite{SUSY-2019-08,hepdata.90607})
with standard Markov-Chain Monte Carlo algorithms.
I recommend first estimating the covariance by expanding about the maximum
\heplikelihood, and linearly transforming your coordinates to whiten that
covariance to the identity matrix.
Then, algorithms provided by
TensorFlow Probability~\cite{tensorflow2015-whitepaper}
(the Metropolis-Adjusted Langevin Algorithm~\cite{mala1998, XIFARA201414}
is a good start)
can sample quite efficiently with tensor evaluation through
\pyhf~\cite{heinrich2021pyhf}.
Histograms of samples for the signal strength in the discovery fits
can differ substantially from profile-\heplikelihood\ curves.
}

>>>>

Despite their troublesome nomenclature, \heplikelihood\ models are extremely
useful and effective for describing the predictions of our hypotheses and
extracting interpretable results.
A general \heplikelihood\ model is a function $H(x)$ that takes input
parameters $x$ and outputs two things:
expected yields $y(x)$ to predict Poisson data,
and an objective value $E$ to be minimized (in a `fit') to choose a best
prediction.
That is,
\begin{equation}
H(x) = [\,y(x), \,E(x)\,]
.
\end{equation}
The minimization objective $E$ is the summed negative $\log$ \heplikelihood\
for all actual and auxiliary data.
We choose the letter $E$ for \textbf{energy}, to highlight the analogy with
statistical physics and the Boltzmann factor
$f(x) \propto e^{-E(x)/(k_B T)}$,
with which a thermodynamic system collapses to its minimum energy at cold
temperatures~\cite{
skilling2017david,
lecun-06,
pmlr-v2-ranzato07a
}.
At temperature $1/k_B$, this can induce a probability distribution
relative to an appropriate measure~\cite{
cranmer2012histfactory,
skilling2017david
},
but in my experience with \heplikelihood\ models such distributions often
coincide with neither intuition nor the best fit value.%
\footnote{%
One can reproduce this experience by sampling a public \pyhf\ \heplikelihood\
model
(for example, I recommend trying SR-HM(disc.) of
$1\ell bb$~\cite{SUSY-2019-08,hepdata.90607})
with standard Markov-Chain Monte Carlo algorithms.
I recommend first estimating the covariance by expanding about the maximum
\heplikelihood, and linearly transforming your coordinates to whiten that
covariance to the identity matrix.
Then, algorithms provided by
TensorFlow Probability~\cite{tensorflow2015-whitepaper}
(the Metropolis-Adjusted Langevin Algorithm~\cite{mala1998, XIFARA201414}
is a good start)
can sample quite efficiently with tensor evaluation through
\pyhf~\cite{heinrich2021pyhf}.
Histograms of samples for the signal strength in the discovery fits
can differ substantially from profile-\heplikelihood\ curves.
}
